# Instrumental Variables

```{r, message=FALSE}
library(dplyr, warn.conflicts = FALSE)
library(stargazer)
library(lmwr)
library(tidyr)
```

## Introduction

## A Confounded Model

### Confounded Model DAG

### Confounded Linear Model

### Simulation of Confounded Data

Here I create a function that returns simulated data as a data frame.

```{r}
set.seed(123456789)

get_data <- function() {
  N <- 1000
  a <- 2
  b <- 3
  c <- 2
  e <- 3
  f <- -1
  d <- 4
  z <- runif(N)
  u_1 <- rnorm(N, mean = 0, sd = 3)
  u_2 <- rnorm(N, mean = 0, sd = 1)
  
  x <- f + d * z + u_2 + c * u_1
  y <- a + b * x + e * u_1
  data.frame(x, y, z)
}

df <- get_data()

lm1 <- lm(y ~ x, data = df)

lm1
```

## IV Estimator

### Graph Algebra of IV Estimator

### Properties of IV Estimator

### IV Estimator with Standard Algebra

### Simulation of an IV Estimator

### IV Esimator with Matrix Algebra

### Two-Stage Least Squares

### IV Estimator in R

```{r}
X <- cbind(1, df$x)
Z <- cbind(1, df$z)
y <- df$y

beta_hat_ols <- solve(t(X) %*% X) %*% t(X) %*% y
beta_hat_iv <- solve(t(Z) %*% X) %*% t(Z) %*% y

beta_hat_ols
beta_hat_iv
```

```{r}
lm_iv <- function(y, X_in, Z_in = X_in, reps = 100,
                  min_in = 0.05, max_in = 0.95) {
  
  set.seed(123456789)
  X <- cbind(1, X_in)
  Z <- cbind(1, Z_in)
  
  bs_mat <- matrix(NA, reps, dim(X)[2])
  
  N <- length(y)
  for (r in 1:reps) {
    # sample(1:N, size = N, replace = TRUE)
    index_bs <- round(runif(N, min = 1, max = N))
    y_bs <- y[index_bs]
    X_bs <- X[index_bs, ]
    Z_bs <- Z[index_bs, ]
    bs_mat[r, ] <- solve(t(Z_bs) %*% X_bs) %*% t(Z_bs) %*% y_bs
  }

  # Present results
  tab_res <- matrix(NA, dim(X)[2], 4)
  tab_res[, 1] <- colMeans(bs_mat)
  for (j in 1:dim(X)[2]) {
    tab_res[j, 2] <- sd(bs_mat[ , j])
    tab_res[j, 3] <- quantile(bs_mat[ , j], min_in)
    tab_res[j, 4] <- quantile(bs_mat[ , j], max_in)
  }
  colnames(tab_res) <- c("coef", "sd", as.character(min_in),
                         as.character(max_in))
  return(tab_res)
  
}

print(lm_iv(df$y, df$x), digits = 3)
print(lm_iv(df$y, df$x, df$z), digits = 3)
```

## Returns to schooling

```{r}
nlsm <-
  nlsm %>%
  mutate(exp = age76 - ed76 - 6,
         exp2 = exp^2/100) %>%
  filter(!is.na(lwage76))

fm <- list()
fm[[1]] <- lm(lwage76 ~ ed76, data = nlsm)
fm[[2]] <- update(fm[[1]], ~ . + exp + exp2)
fm[[3]] <- update(fm[[2]], ~ . + black + reg76r)
fm[[4]] <- update(fm[[3]], ~ . + smsa76r + smsa66r + reg662 + reg663 +
                    reg664 + reg665 + reg666 + reg667 + reg668 + reg669)
fm[[5]] <- update(fm[[4]], ~ . - ed76 + nearc4)
fm[[6]] <- update(fm[[5]], ed76 ~ .)
```

```{r, results="asis"}
stargazer(fm, type = "html", keep.stat = c("n", "rsq"),
          omit = "^(reg66|smsa)")
```

```{r}
fm[[5]]$coefficients[2]/fm[[6]]$coefficients[2]
fm[[5]]$coefficients[["nearc4"]]/fm[[6]]$coefficients[["nearc4"]]

iv_fit <- function(y_in, X_in, Z_in = X_in) {
  # Convert supplied data to matrices
  y <- as.matrix(y_in, ncol=1)
  X <- cbind(1, as.matrix(X_in))
  Z <- cbind(1, as.matrix(Z_in))

  res <- solve(t(Z) %*% X) %*% t(Z) %*% y
  rownames(res) <- c("intercept", colnames(X_in))
  return(res)
}

iv_bs <- function(i, y, X_in, Z_in) {

  y_vec <- as.matrix(y, ncol=1)
  N <- length(y_vec)

  index_bs <- round(runif(N, min = 1, max = N))
  y_bs <- y_vec[index_bs, ]
  X_bs <- X_in[index_bs, ]
  Z_bs <- Z_in[index_bs, ]

  as_tibble(t(iv_fit(y_bs, X_bs, Z_bs)))
}

lm_iv <- function(y, X_in, Z_in = X_in, reps = 100) {

  set.seed(123456789)
  bind_rows(lapply(1:reps, iv_bs, y, X_in, Z_in))
}

get_summ_stats <- function(x, p_lower, p_upper) {
  df_temp <- tibble(mean(x),
                    sd(x),
                    quantile(x, p_lower),
                    quantile(x, p_upper))

  names(df_temp) <- c("coef", "sd",
                         paste0("p", as.character(p_lower * 100)),
                         paste0("p", as.character(p_upper * 100)))
  return(df_temp)
}

get_tab_res <- function(bs_mat, p_lower = 0.05, p_upper = 0.95) {
  bind_rows(lapply(bs_mat, get_summ_stats, p_lower, p_upper),
            .id = "var")
}

y <-
  nlsm %>%
  select(lwage76)

X <-
  nlsm %>%
  select(ed76, exp, exp2, black, reg76r,
                     smsa76r, smsa66r, reg662:reg669)
Z1 <-
  nlsm %>%
  mutate(age2 = age76^2) %>%
  select(nearc4, age76, age2, black, reg76r,
         smsa76r, smsa66r, reg662:reg669)

Z2 <-
  nlsm %>%
  mutate(age2 = age76^2) %>%
  select(momdad14, age76, age2, black, reg76r,
         smsa76r, smsa66r, reg662:reg669)

res_Z1 <- lm_iv(y, X, Z1, reps = 1000)
get_tab_res(res_Z1)

res_Z2 <- lm_iv(y, X, Z2, reps = 1000)
bs_diff <- (res_Z1 - res_Z2)[, "ed76"]
summary(bs_diff)

# 3.4.5 Concerns with Distance to College
summarise_by <- function(.data, .by, ..., .groups = NULL) {
  .data %>%
    summarise(across(c(ed76, exp, black, south66,
                       smsa66r, reg76r, smsa76r), mean),
              .groups = "keep") %>%
    pivot_longer(cols = everything())
    pivot_wider(names_from = .by, values_from = value)
}

nlsm %>%
  mutate(nearc4 = case_when(nearc4 == 0 ~ "Not near college",
                            nearc4 == 1 ~ "Near college")) %>%
  group_by(nearc4) %>%
  summarise(across(c(ed76, exp, black, south66,
                     smsa66r, reg76r, smsa76r), mean))

```
