# Panel Data

## First differences

### First difference model

### Simulated panel data

```{r, message=FALSE}
library(dplyr, warn.conflicts = FALSE)
library(lmwr)
library(tidyr)
library(stargazer)
library(lfe) 
```

```{r}
set.seed(123456789)
N <- 1000
T <- 2
a <- -2
b <- 3
d <- -4
u <- c(5, 6)
```

```{r}
x1 <- runif(N)
x2 <- x1 + runif(N)
x <- as.matrix(rbind(x1, x2))
e <- matrix(rnorm(N * T), nrow = T)
y <- a + b * x + d * u + e
```

```{r}
diffy <- y[2, ] - y[1, ]
diffx <- x[2, ] - x[1, ]
lm1 <- lm(diffy ~ diffx)
```

```{r, results='asis', echo=FALSE}
library(xtable)
print(xtable(summary(lm1)$coefficients), type="html")
```

If we wanted data in a tidy data frame, we could use functions from `tidyr` to rearrange the matrices and vectors above into a tidy data frame with and `id` column for each individual and a `time` column for periods $1$ and $2$.
However, as the following code suggests, while the code works, this approach is quite unwieldy.


```{r}
df <- 
  tibble(id = 1:(dim(y)[2]), y_1 = y[1,], y_2 = y[2, ], 
             x_1 = x[1, ], x_2 = x[2,]) %>% 
  pivot_longer(cols = y_1:x_2) %>%
  separate(name, into = c("var", "time"), sep = "_", 
           convert = TRUE) %>%
  pivot_wider(id_cols = c("id", "time"), names_from = "var",
              values_from = "value")

df
```

A better approach is probably to build up the data using a "tidy" approach from the outset.
Note that the `arrange` below is only needed to ensure that the results line up with those above and could be omitted in general.

```{r}
set.seed(123456789)

df_pre <- tibble(id = 1:N, time = 1L, x = runif(N), u = 5)
df_post <- tibble(id = 1:N, time = 2L, x = df_pre$x + runif(N), u = 6)

df_all <-
  bind_rows(df_pre, df_post) %>%
  arrange(id, time) %>%
  mutate(e = rnorm(N * T)) %>%
  mutate(y = a + b * x + d * u + e) %>%
  select(-u, -e)

df_all
```

### OLS estimation of first differences

```{r}
df %>%
  group_by(id) %>%
  arrange(time) %>%
  mutate(diffy = y - lag(y),
         diffx = x - lag(x)) %>%
  lm(diffy ~ diffx, data = .)
```

## Difference in difference

### Difference-in-difference estimator

### Difference-in-difference estimator in R

The following version of the `f_did` function substracts four lines from that in the book.

```{r}
f_did <- function(y, treat) {
  y1 <- y[1, ]
  y2 <- y[2, ]
  did <- matrix(NA, 3, 3)
  did[1, 1] <- mean(y2[treat==1])
  did[2, 1] <- mean(y1[treat==1])
  did[1, 2] <- mean(y2[treat==0])
  did[2, 2] <- mean(y1[treat==0])
  did[3, ] <- did[1, ] - did[2, ]
  did[ , 3] <- did[ , 1] - did[ , 2]
  row.names(did) <- c("Period 2", "Period 1", "Diff")
  colnames(did) <- c("Treated", "Not Treated", "Diff")
  return(did)
}
  
```

```{r}
set.seed(123456789)
treat <- runif(N) < 0.5
x2 <- x1 + treat
x <- rbind(x1, x2)
y <- a + b * x + d * u + e
did1 <- f_did(y, treat)
```

```{r, results='asis', echo=FALSE}
print(xtable(did1), type="html")
```

## Minimum wage increase in New Jersey

Here I was too lazy to do the "base R" version.
Note that Adams (2021) states "Card and Krueger (1994) present results on a measure called full-time equivalence.
It is unclear how that measure is calculated."
Actually if you go to the website for the data, you can see from the provided SAS code that it is calculated as below.
I don't quite reproduce the results in Table 3 of [the paper](https://davidcard.berkeley.edu/papers/njmin-aer.pdf), but I get very close.

### Data from Card and Krueger (1994)

We used `cardkrueger` from `lmwr`.

```{r}
nj_min_pre <- 
  cardkrueger %>%
  select(sheet, empft, emppt, state, nmgrs) %>%
  rename(ft = empft, pt = emppt) %>%
  mutate(time = 1) 

nj_min_post <- 
  cardkrueger %>%
  select(sheet, empft2, emppt2, state, nmgrs2) %>%
  rename(ft = empft2, pt = emppt2, nmgrs = nmgrs2) %>%
  mutate(time = 2)

nj_min <- 
  bind_rows(nj_min_pre, nj_min_post) %>%
  mutate(total = ft + pt,
         fte = ft + pt * 0.5 + nmgrs,
         post = time == 2, 
         treat = state == 1) %>%
  group_by(sheet) %>% 
  mutate(balanced = sum(!is.na(total)) == 2) %>%
  ungroup() 
```

### Difference-in-Difference Estimates

```{r}
fms <- list(lm(total ~ post * treat, data = nj_min),
            lm(total ~ post * treat, data = nj_min, subset = balanced),
            lm(fte ~ post * treat, data = nj_min),
            lm(fte ~ post * treat, data = nj_min, subset = balanced))
```

```{r, results="asis", echo=FALSE}
stargazer(fms, keep.stat = c("n", "rsq"), type = "html")
```

## Fixed Effects

### Fixed-Effects Estimator

### Nuisance Parameter

### Adjusted Fixed-Effects Estimator

### Two-Step Fixed-Effects Estimator

### Fixed Effects Estimator in R

Note that the code in the book incorrectly describes `XT` and `XC` ("in each row" and "in each column" are flipped).

```{r}
f_fe <- function(Y, X = NULL, cross = TRUE) {
  Y <- as.matrix(Y)
  T <- dim(Y)[1]
  N <- dim(Y)[2]
  
  # Creates a T x N matrix with numbers 1 to T in each column
  XT <- matrix(rep(c(1:T), N), nrow = T)
  y <- as.vector(Y)
  t <- as.vector(XT)
  
  # Set up for different cases
  if (cross) {
    # Create cross-section dummies
    
    # Creates a T x N matrix with 1 to N in each row
    XC <- t(matrix(rep(c(1:N), T), nrow = N))
    c <- as.vector(XC)
  }
  
  if (!is.null(X)) {
    # Create treatment variable
    X <- as.matrix(X)
    treat <- as.vector(X)
  }
  
  # Estimator
  if (cross & !is.null(X)) {
    # Standard case
    lm1 <- lm(y ~ treat + as.factor(t) + as.factor(c))
  } else if (!is.null(X)) {
    # No cross-section
    lm1 <- lm(y ~ treat + as.factor(t))
  } else {
    # No treatment
    lm1 <- lm(y ~ as.factor(t))
  }
  return(lm1)
}
```

The simulated panel data set has 100 individuals observed over 10 time periods.

```{r}
set.seed(123456789)
N <- 100
T <- 10
alpha <- runif(N)
gamma <- runif(T)
beta <- 3
epsilon <- matrix(rnorm(N * T), nrow = T)
treat <- runif(N) < 0.5
y <- t(matrix(rep(alpha, T), nrow = N)) + gamma + epsilon
y[1, ] <- y[1, ] + beta * treat
treat1 <- matrix(0, T, N)
treat1[1, ] <- treat
```

```{r}
# Standard estimator
lm1 <- f_fe(y, treat1)

# No invidividual fixed effects
lm2 <- f_fe(y, treat1, cross = FALSE)

# Adjusted estimator
y0 <- y[2:T, ] 
alpha_hat <- colMeans(y0) # Calculate alpha
y2 <- y - t(matrix(rep(alpha_hat, T), nrow = N))
lm3 <- f_fe(y2, treat1, cross = FALSE)

# Two-step estimator
lm4 <- f_fe(y0, cross = FALSE)
y0_res <- matrix(lm4$residuals, nrow = T - 1)
alpha_hat <- colMeans(y0_res) # Calculate alpha
y3 <- y - t(matrix(rep(alpha_hat, T), nrow = N)) # Adjust outcome
lm5 <- f_fe(y3, treat1, cross = FALSE)
```

```{r, results="asis"}
stargazer(list(lm1, lm2, lm3, lm5),
          keep.stat = c("n", "rsq"), keep = "treat", type = "html")
```

Now, rearrange the data into a "tidy" form.

```{r}
df <-
  tibble(y = as.vector(y), 
         treat = as.vector(treat1),
         id = as.vector(t(matrix(rep(c(1:N), T), nrow = N))),
         t = as.vector(rep(c(T:1), N)))

df %>% print(n = 15)
```

And use the `felm` function from the `lfe` package to reproduce the regression results above.

```{r} 
felm1 <- felm(y ~ treat | id + t, data = df)
felm2 <- felm(y ~ treat | t, data = df)

df_alpha_hat <-
  df %>% 
  filter(t < 10) %>%
  mutate(y0_res = residuals(lm(y ~ as.factor(t), data = .))) %>%
  group_by(id) %>%
  summarize(alpha_hat = mean(y0_res),
            mean_y = mean(y), .groups = "drop")

felm3 <-
  df %>%
  inner_join(df_alpha_hat, by = "id") %>%
  mutate(y = y - mean_y) %>%
  felm(y ~ treat | t, data = .)
  
felm5 <-
  df %>%
  inner_join(df_alpha_hat, by = "id") %>%
  mutate(y = y - alpha_hat) %>%
  felm(y ~ treat | t, data = .)
```   

```{r, results="asis"}
stargazer(list(felm1, felm2, felm3, felm5),
          keep.stat = c("n", "rsq"), keep = "treat", type = "html")
```
